Using structural and evolutionary information to detect and correct pyrosequencing errors in non-coding RNAs

### 2013-05-28 JCB/results_JCB.tex
Rephrased 16s

\subsection{Error Correction in 5s rRNAs}
\label{sec:5S}
To illustrate the potential of our algorithm, we applied our techniques to identify and correct point-wise errors in RNA sequences
with conserved secondary structures. More precisely, we used \RNApyro to reconstruct 5s rRNA sequences with randomly distributed
mutations. This experiment has been designed to suggest further applications to error-corrections in pyrosequencing data.

@@

\subsection{Error Correction in 16s rRNAs}
\label{sec:16S}
\TODOTous{Describe and present results????}

To complete our benchmark, we considered the small subunit ribosomal RNA, a molecule wich is of particular interest  in metagenomics an phylogenetics.
We gathered the seed sequences of the bacterial multiple sequence alignment (MSA) retrieved from the RFAM Database 11.0 (Rfam id: RF00177)~\cite{gardner2011rfam}.
This alignment is composed of 93 sequences, ranging between 1461 and 1568 nucleotides long, and yield an average pairwise sequence similarity of $69\%$.

As in Sec.~\ref{sec:5S} we used a leave-one-out strategy. Since few structures contain all gaps 
at the same positions, after selecting a sequence of interest,  we used the following strategy.
 It was applied to each sequence in the alignment. Starting with the full alignment, all positions corresponding to gaps in the sequence of interest were removed, for all sequences in the alignment.

To simulate sequencing error in the sequence of interest,
 we use the the next-generation sequencing read simulator 
ART~\cite{huang2012art}. The Illumina technology main error mode is base substitution. We used
this model since those
are the errors that \RNApyro detects. As parameters, the reads  generated
were $75\text{bps}$ long for an average fold coverage of $5$, $10$ and $15$. 
 They yield respectively an average error rate of $2.4\%$, $0.9\%$ and $0.6\%$.

We used a value of $\beta=15$ and evaluated $\alpha$ at $0, 0.5, 0.8, 1.0$. 
 Additionally since the alignment still contains gap, we made sure to evaluate the isostericity 
 term only
 with sequences in the MSA where both positions in the base pair are not gaps. 
 Due to the low number of errors, we set the presumed as twice the average error rate made by ART, i.e. $4.8\%$, $1.8\%$ and $1.2\%$ for fold coverages of $5$, $10$ and $15$ respectively.

To evaluate our method, we simulated with a random sequences from the MSA every time,
 $12$ times the $5-$fold, $10$ times the $10-$fold and $14$ times the $15-$fold coverage. And we 
 evaluated the ROC curve mostly as in Sec.~\ref{sec:5S}. In Fig.~\ref{fig:16s_all} they are computed
 over all positions, while in Fig.~\ref{fig:16s_bp} only over positions inside a base pair.

@@

Our data demonstrates that even on long sequences our algorithm shows interesting potential for 
error-correction applications. First, the AUC values (up to 0.81) when looking at all positions, 
in Fig.~\ref{fig:16s_all},  indicate that a signal has been successfully extracted. When restraining
to structured regions (i.e. base paired regions) we obtain AUC values up to 0.998.  Since 
contributions to the energy only comes from structured regions, this was an expected result.

An interesting feature is that best results are almost always when $\alpha=0$, i.e. all the contribution
comes from the isostericity. A notable exception is when the number of mutation is underestimated, 
in that case best performances were observed for non-nil values of $\alpha$.

>>>>

\subsection{Homogenous Error Correction in 5s rRNAs}
\label{sec:5S}
To illustrate the potential of our algorithm, we applied our techniques to identify and correct point-wise errors in RNA sequences
with conserved secondary structures. More precisely, we used \RNApyro to reconstruct 5s rRNA sequences with randomly distributed
mutations. This experiment has been designed to suggest further applications to error-corrections in pyrosequencing data.

@@

\subsection{Correcting Illumina sequencing errors in 16s rRNAs}
\label{sec:16S}
\TODOTous{Describe and present results????}

To complete our benchmark, we turned to the small subunit ribosomal RNA in bacteriae, a molecule which is of particular interest  in metagenomics and phylogenetics. Our aim was to get as close as possible to a pyrosequencing context, in which reads are produced non-uniformly by an Illumina sequencer, impacting the distribution of errors in the sequence. 
We chose this setting both because of the popularity of Illumina sequencing in metagenomics, and since the underlying sequencing technique only considers base substitutions (no insertions), the only type of errors currently detected by \RNApyro.
To that purpose, we used simulated Illumina reads, mapped the reads back to a reference alignment, and run \RNApyro on a consensus sequence derived from the mapped reads, estimating the maximal amount of mutations from both the length of the sequence and the sequencing depth.

We gathered the seed sequences of the bacterial multiple sequence alignment (MSA) retrieved from the RFAM Database 11.0 (Rfam id: RF00177)~\cite{gardner2011rfam}. This alignment is composed of 93 sequences, whose length varies between 1461 and 1568 nucleotides, and has an average pairwise sequence identity of $69\%$. We used the pseudoknot-free version of the consensus secondary structure. A secondary structure for a specific reference sequence was obtained by simply mapping the structure back to sequence, i.e. by removing any base-pair having at least one partner involved in a gapped position. For similar reasons, we locally excluded from our calculation of the isostericity contribution for a given base-pair, described in Section~\ref{sec:model}, any sequence featuring at least a gap on the corresponding positions.

To simulate sequencing errors, we used the next-generation sequencing read simulator 
{\tt ART}~\cite{huang2012art}. The \emph{Illumina technology} setting was chosen as the main error mode, generating reads of $75~\text{bps}$, featuring mostly base substitution errors. Reads were mapped back to the original sequence, and a consensus sequence was determined, from the sequencing output, by a simple majority vote in the case of multiple coverage. Uncovered regions were simply generated at random. The average  rate of errors observed in the final consensus sequence was empirically estimated to represent $2.4\%$, $0.9\%$ and $0.6\%$ of the reference sequence, for prescribed sequencing coverages of $5$, $10$ and $15$ fold respectively.

As in Section~\ref{sec:5S}, we evaluated the predictive power of \RNApyro-computed mutational profiles using a leave-one-out strategy. We picked a sequence at random from the MSA, sequenced/mutated it as above, and run \RNApyro to establish its mutational profile. In this execution, we used values of $\beta=15$ and $\alpha\in [0, 0.5, 0.8, 1.0]$, and set the presumed number of mutations to twice the average error rate made by {\tt ART}, i.e. $4.8\%$, $1.8\%$ and $1.2\%$ for fold coverages of $5$, $10$ and $15$ respectively.
We repeated the whole procedure $12$/$10$/$14$ times for the $5$/$10$/$15-$fold coverage. 
We evaluated the predictive power of the profile by a computing a joint ROC curve for each value of $\alpha$, and each coverage, as described in Section~\ref{sec:5S}. Figure~\ref{fig:16s_all} shows the ROC curves, computed over all positions, while Figure~\ref{fig:16s_bp} only focuses on positions that are paired in the consensus secondary structure.

@@

Our data demonstrates that even on long sequences our algorithm shows interesting potential for 
error-correction applications. First, the AUC values (up to 0.81) when looking at all positions, 
in Fig.~\ref{fig:16s_all},  indicate that a signal has been successfully extracted. When restraining
to structured regions (i.e. base paired regions) we obtain AUC values up to 0.998.  Since 
contributions to the energy and isostericity only arise from structured regions, this was an expected result.

An interesting feature is that best results are almost always when $\alpha=0$, i.e. all the contribution
comes from the isostericity. A notable exception is when the number of mutation is underestimated, 
in that case best performances were observed for non-null values of $\alpha$.
